---
title: "Higher education % in Oregon - Alis volat propriis"
output:
  html_document:
    df_print: paged
---
#### Welcome to my notebook covering the use of the tidycensus R package for extracting and evaluating census data in the state of Oregon. In part A We are interested in uncovering the percentages of the population with graduate level degrees at a county level. Part B will continue to focus on education data, Stay tuned to find out more!

### Import Libraries

##### Start by importing the list of libraries that will assist us in connecting with the Census Bureau's APIs and visualizing the desired data. 
```{css, echo=FALSE}
p {
  margin-bottom: 0px;
}
```
```{r message = FALSE, warning = FALSE}
library(tidycensus) # connect with the census bureau API
library(tidyverse) # data wrangling
library(mapview) # spatial data visualizer
library(plotly) # graph / chart visualizer
library(ggiraph) # graph / chart visualizer
library(ggplot2) # graph / chart visualizer
library(DT) # fancy tables
library(scales) # scales map data to aesthetics
library(sf) # spatial data package
```
### Part A
<hr />
### Data Source

##### We will use the get_acs() function from the tidycensus package to pull data from the census bureau. The function defaults to the 2017-2021 5-year ACS dataset. Which is necessary for us, since there are a list of counties in Oregon with a population of less than 65,000. 

##### The data we would like to visualize is the percent of the population that have a graduate degree at the county level in the state of Oregon.  
```{r message = FALSE, warning = FALSE}
# use the get_acs() function for importing data
or_grads <- get_acs(
  geography = "county", # county level geography
  variables = "DP02_0066P", # variable for percent of the population that have a graduate degrees
  state = "OR", 
  year = 2021
)
```
### Data Manipulation & DataTables

##### An important aspect of the tidycensus package is that is presents the data to us in a tidy format. The data is pre-cleaned and takes little wrangling to answer questions.
```{r}
# filter and arrange results
or_grads_table <- or_grads %>%
  separate(NAME, into = c("county", "state"), sep = ", ") %>% # create county and state columns
  arrange(-estimate) %>% # sort by estimate descending
  select(county, state, estimate, moe) # select important columns
```

##### By using the datatable() function from the DT library we can filter our results to see the desired information.
```{r}
# function from the DT library to create a datatable
datatable(or_grads_table)
```
**Figure 1** above reveals the data imported from the get_acs() function. Results conclude that the top 3 counties with graduate degrees is Benton, Multnomah, and Washington. Bottom 3 counties are Malheur, Gilliam, and Morrow.

<hr />

### Data Visualization & Analysis

##### Now that our data is imported and in a proper format. We can begin using ggplot to visualize the percentage of the oregon population that graduate have degrees. 
```{r fig.height=6, fig.width=9.5}
# use ggplot with estimate as the x axis and county as the y axis
or_plot <- ggplot(or_grads_table, aes(x = estimate, 
                                      y = reorder(county, estimate))) + # sort by counties high to low
  geom_point(color="navy", size = 3) + # state color navy blue for point data
  scale_x_continuous(labels = function(x) paste0(x, '%')) + # concatenate '%' to estimate data
  scale_y_discrete(labels = function(x) str_remove(x, " County")) + # strip 'county' from the name column
  labs(title = "% Population with graduate degrees, 2017-2021 ACS", # use appropriate title
       subtitle = "Counties in Oregon", # use appropriate subtitle
       caption = "Data acquired with R and tidycensus",
       x = "ACS estimate Percentage",
       y = "") +
  theme_minimal(base_size = 12) # remove tick marks 
  
or_plot # print plot
```
**Figure 2** above plots the counties by estimate value in a descending order. We get a full picture of our results without having to filter through multiple pages in a table.

< hr />

##### All tables in the tidycensus package come with a margin of error column to give us an understanding of accuracy in the dataset. Typically, census data from geographies of larger population have a lower margin of error.

##### Lets plot the MOE variable with our estimate variable below. 
```{r fig.height=6, fig.width=9.5}
# create plot revealing moe and estimate variables
or_plot_errorbar <- ggplot(or_grads, aes(x = estimate, y = reorder(NAME, estimate))) + # sort by counties high to low
  geom_errorbar(aes(xmin = estimate - moe, xmax = estimate + moe), 
                width = 0.5, linewidth = 0.5) +
  geom_point(color = "gold", size = 3) + # state color gold for point data
  scale_x_continuous(labels = function(x) paste0(x, '%')) + # concatenate '%' to estimate data
  scale_y_discrete(labels = function(x) str_remove(x, " County, Oregon|, Oregon")) + # strip 'county' from the name column
  labs(title = "% Population with graduate degrees, 2017-2021 ACS", # use appropriate title
       subtitle = "Counties in Oregon", # use appropriate subtitle
       caption = "Data acquired with R and tidycensus. Error bars represent margin of error around estimates.",
       "ACS estimate",
      y = "",
      x = "Percentage") +
  theme_minimal(base_size = 12) # remove tick marks 
  
or_plot_errorbar # print plot
```
**Figure 3** above reveals the margin of error associated with the estimate data.

<hr />

### Interative Plots

##### Using ggplotly() I'm going to recreate the plot above but leveraging the interactivity of the plotly library.
```{r fig.height=6, fig.width=9.5}
# use ggplotly() and add a tooltip, and extend the margins of the plot
ggplotly(or_plot_errorbar, tooltip = "x")%>% 
  layout(margin = list(l = 50, r = 50, b = 50, t = 50))
```
**Figure 4** Hover over the points in the plot above to see the interactive tootip.

<hr />

### Part B

<hr />

##### Continuing with the theme of education, lets import more data through the tidycensus package. 

**What percentage of people with bachelor's degrees rent or own their home in the state of Oregon at the tract level?**

##### To find variables that will assist us in our analysis, we will use the load_variables() function to filter through the ACS data. 
```{r message = FALSE, warning = FALSE}
# use the load_variables function() with the year and ACS 5 year parameter
vars <- load_variables(2021, "acs5")
# load vars into a datatable
datatable(vars)
```
**Figure 5** By using the search feature in the datatable, you see examples of available results. For example search "bachelor's" to find variable codes moving forward. 

<hr />

### Import Data

##### We will need to pull in two datasets and eventually merge the results into one table. 
##### -- "B25013_006" is the population of people with bachelor's degrees that own their home.
##### -- "B25013_011" is the population of people with bachelor's degrees that rent their home.
```{r message = FALSE, warning = FALSE}
# import data at the tract level for home owners
bach_owners <- get_acs(
  geography = "tract",
  variables = "B25013_006",
  state = "OR",
  geometry = TRUE # get multipolygon data of Oregon tracts
)

# import data at the tract level for renters
bach_renters <- get_acs(
  geography = "tract",
  variables = "B25013_011",
  state = "OR",
  geometry = FALSE # geometry is false because we are getting our multipolygon data in the owners dataset. 
)
```
### Data Wrangling

##### Use functions of the tidyverse to rename, select, and mutuate columns to fit our analysis. 
```{r}
# clarify the name of the estimate & moe columns before joining
bach_owners <- bach_owners %>%
  rename(estimate_owners = estimate, moe_owners = moe)

# clarify the name of the estimate & moe columns before joining
bach_renters <- bach_renters %>%
  rename(estimate_renters = estimate, moe_renters = moe) %>%
  select(GEOID, estimate_renters, moe_renters) # drop uneccessary values

# join dataframes on GEOID, and create percent columns of each category
bach_housing <- left_join(bach_owners, bach_renters, by='GEOID') %>%
  select(GEOID, NAME, estimate_owners, estimate_renters, moe_owners, moe_renters, geometry) %>%
  mutate(estimate_total = estimate_renters + estimate_owners) %>%
  mutate(percent_renters = (estimate_renters/estimate_total)*100) %>%
  mutate(percent_owners = (estimate_owners/estimate_total)*100)
```


```{r}
mapview(bach_housing, zcol = "percent_renters", zcol = "percent_owners")
```